---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# a5R

<!-- badges: start -->
[![R-CMD-check](https://github.com/belian-earth/a5R/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/belian-earth/a5R/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/belian-earth/a5R/graph/badge.svg)](https://app.codecov.io/gh/belian-earth/a5R)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![extendr](https://img.shields.io/badge/extendr-^0.8.1-276DC2)](https://extendr.github.io/extendr/extendr_api/)
<!-- badges: end -->

a5R provides R bindings for the [A5](https://a5geo.org/) pentagonal discrete
global grid system (DGGS), powered by the
[a5 Rust crate](https://crates.io/crates/A5) via
[extendr](https://extendr.github.io/extendr/extendr_api/).

a5 partitions the Earth's surface into pentagonal cells across 31 resolution
levels. Cells are equal-area, encoded as 64-bit integers, and achieve
millimetre-level precision at the finest resolution. For a full description of
the system, see the [A5 project page](https://a5geo.org/) and the
[reference implementation](https://github.com/felixpalmer/a5) by Felix Palmer.

## R-specific design

- **vctrs**: Cell indices are represented as an `a5_cell` vector type built on
  [vctrs](https://vctrs.r-lib.org/), giving you type safety, column support in
  tibbles, and natural coercion to/from character.
- **wk**: Boundary geometries and cell centroids are returned as
  [wk](https://paleolimbot.github.io/wk/) geometry vectors (`wk_wkt` and `xy`),
  so they integrate directly with sf, terra, and other spatial tooling.

## Installation

You can install the development version of a5R from
[GitHub](https://github.com/belian-earth/a5R) with:

``` r
# install.packages("pak")
pak::pak("belian-earth/a5R")
```

You will need a working [Rust toolchain](https://rustup.rs/) (`cargo` and
`rustc`).

## Examples

```{r example}
library(a5R)
```

Index a point to a cell at resolution 5:

```{r}
cell <- a5_lonlat_to_cell(-3.19, 55.95, resolution = 5)
cell
```

Convert back to coordinates:

```{r}
a5_cell_to_lonlat(cell)
```

Get the cell boundary polygon:

```{r}
a5_cell_to_boundary(cell)
```

Navigate the hierarchy:

```{r}
a5_cell_to_parent(cell)
a5_cell_to_children(cell)
```

Cell info:

```{r}
a5_get_resolution(cell)
a5_cell_area(0:5)
```

### Visualising the grid hierarchy

```{r hierarchy-plot, fig.alt = "Pentagonal A5 grid cells at two resolutions nested inside each other"}
# A cell and its children two levels down
parent <- a5_lonlat_to_cell(0, 0, resolution = 3)
children <- a5_cell_to_children(parent, resolution = 5)

# wk geometries plot directly with base graphics
plot(a5_cell_to_boundary(children), col = "#206ead20", border = "#206ead", asp = 1)
plot(a5_cell_to_boundary(parent), border = "#333333", lwd = 2, add = TRUE)
```


### Grid generation

Generate all cells covering an area with `a5_grid()`:

```{r grid-plot, fig.alt = "A5 grid cells covering Edinburgh at resolution 8"}
cells <- a5_grid(c(114.8, 4.1, 119.8, 8.1), resolution = 8)
plot(a5_cell_to_boundary(cells), col = "#206ead20", border = "#206ead", asp = 1)
```

Any geometry that [wk](https://paleolimbot.github.io/wk/) can handle works
as input, including sf objects and WKT strings. Bounding boxes that cross the
antimeridian are supported too:

```{r}
# Fiji — bbox crosses the antimeridian
a5_grid(c(177, -19, -178, -17), resolution = 5)
```


## Performance

a5R provides a performant interface to the A5 indexing system, with all
core operations implemented in Rust and called directly from R via
[extendr](https://extendr.github.io/). Cross-language
[benchmarks](https://github.com/belian-earth/a5R/blob/main/benchmarks/RESULTS.md)
and consistency checks against the Python, JavaScript, and DuckDB A5
implementations confirm that results are identical across all APIs.

## Acknowledgements

A5 was created by [Felix Palmer](https://github.com/felixpalmer). This package
is a thin R wrapper around his work and would not exist without it. The [Query-farm](https://github.com/Query-farm/a5) team maintain the
DuckDB A5 extension, which wraps the same Rust crate and provided a valuable reference for this project.