---
title: "Getting started with a5R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with a5R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(a5R)
```

## Index a point

Map a longitude/latitude coordinate to a cell at a given resolution (0--30).
Higher resolutions produce smaller cells.

```{r}
cell <- a5_lonlat_to_cell(-3.19, 55.95, resolution = 10)
cell
```

Convert back to the cell centre point:

```{r}
a5_cell_to_lonlat(cell)
```

## Cell boundaries

Get the boundary polygon for one or more cells:

```{r boundary-plot, fig.width = 5, fig.height = 4}
boundary <- a5_cell_to_boundary(cell)
boundary

plot(boundary, col = "#206ead20", border = "#206ead", asp = 1)
```

Boundaries are returned as `wk_wkb` vectors by default (set `format = "wkt"` for
WKT). Both integrate directly with sf, terra, and other spatial tooling via the
wk package.

## Hierarchy and compaction

A5 is a hierarchical grid: every cell has a **parent** at a coarser
resolution and 4 **children** at the next finer resolution.

```{r}
parent <- a5_cell_to_parent(cell)
parent

children <- a5_cell_to_children(cell)
children
```

We can visualise the relationship: the parent (dark outline) contains
our cell (blue fill), which in turn contains its 4 children (orange):

```{r hierarchy-plot, fig.width = 8, fig.height = 8}
plot(NULL, xlim = c(-3.23, -3),  ylim = c(55.98, 55.99),               
       xlab = "", ylab = "", asp = 1
       )
plot(a5_cell_to_boundary(a5_cell_to_children(cell)),
     col = "#ad6e2020", border = "#ad6e20", add = TRUE)
plot(a5_cell_to_boundary(cell), col = "#206ead40", border = "#206ead",
     lwd = 2, add = TRUE)
plot(a5_cell_to_boundary(parent), border = "#333333", lwd = 2, add = TRUE)
```

Cell area decreases geometrically: each level is roughly 4x smaller.

```{r}
a5_cell_area(0:5)
```

### Compact and uncompact

When a complete set of siblings is present, `a5_compact()` merges them
back into their shared parent. This is the inverse of `a5_cell_to_children()`
and is useful for reducing the size of large cell sets without losing
coverage.

```{r}
children
a5_compact(children)

# round-trips back to the original
a5_uncompact(a5_compact(children), resolution = 11)
```

Many a5R functions return compacted output automatically. For example,
`a5_grid_disk()` and `a5_spherical_cap()` compact their results; use
`a5_uncompact()` when you need a uniform-resolution grid (see
[Traversal](#traversal) below).


## Traversal

Find neighbouring cells by hop count with `a5_grid_disk()`, or by
great-circle distance with `a5_spherical_cap()`:

```{r traversal, fig.width = 7}
disk <- a5_grid_disk(cell, k = 10)
cap <- a5_spherical_cap(cell, radius = 50000)

plot(a5_cell_to_boundary(cap), col = "#6ead2020", border = "#6ead20", asp = 1)
plot(a5_cell_to_boundary(disk), col = "#206ead20", border = "#206ead", asp = 1)
```

Both functions return a **compacted** cell vector: sibling groups are
merged into coarser parent cells to save space. To recover a uniform grid
at the original resolution, pass the result through `a5_uncompact()`:

```{r traversal-uncompact, fig.width = 7}
disk_grid <- a5_uncompact(disk, resolution = a5_get_resolution(cell))

plot(a5_cell_to_boundary(disk_grid), col = "#206ead20", border = "#206ead", asp = 1)
```


## Converting geometries to a5 cells

`a5_polygon_to_cells()` returns the cells whose centres lie inside a
polygon. It accepts any geometry that wk can handle: a `wk::rct()`
bounding box, a WKT or WKB polygon, or an `sf` / `sfc` feature.

```{r grid-plot, fig.width = 7}
cells <- a5_polygon_to_cells(wk::rct(-3.3, 55.9, -3.1, 56.0), resolution = 12)
length(cells)

plot(a5_cell_to_boundary(cells), col = "#206ead20", border = "#206ead", asp = 1)
```

The returned vector is sorted and compacted: whenever four sibling
cells all sit inside the polygon, they are merged into their parent so
the result uses fewer slots without losing coverage. Call
`a5_uncompact()` to expand back to a uniform grid at the target
resolution:

```{r}
cells_uncom <- a5_uncompact(cells, 12)
length(cells_uncom)

plot(a5_cell_to_boundary(cells_uncom), col = "#206ead20", border = "#206ead", asp = 1)
```

Multi-part inputs are handled natively: a `MULTIPOLYGON` or an `sfc`
of multiple polygons returns the union of cells across all parts, and
a `POLYGON` with holes returns the outer ring's cells with the hole
cells properly subtracted.

```{r, eval = requireNamespace("sf", quietly = TRUE)}
library(sf)
demo(nc, ask = FALSE, echo = FALSE)
nca5 <- a5_polygon_to_cells(nc, resolution = 9) |>
  a5_uncompact(9)
plot(a5_cell_to_boundary(nca5), col = "#03030320", border = "#6d20adff", asp = 1)
```

## Tracing a line

`a5_linestring_to_cells()` returns the cells whose pentagons are
crossed by a great-circle polyline. Output is in discovery order along
the path. Consecutive waypoints are connected by great-circle arcs, so
antimeridian-crossing routes work transparently.

Multi-part inputs (`MULTILINESTRING` or an `sfc` of several linestrings)
are handled natively: per-feature outputs are concatenated in feature
order with first-seen deduplication.

To show what cell-by-cell tracing looks like, we can write a short
phrase as a `MULTILINESTRING` (one stroke per letter) and ask A5 to
fill in the cells:

```{r linestring-a5R, fig.width = 7, fig.height = 3}
a5R_strokes <- wk::wkt(
  "MULTILINESTRING (
    (1.8 52.5, 1.2 52.8, 0.6 52.5, 0.4 52, 0.6 51.5, 1.2 51.2, 1.8 51.5, 2 52, 1.8 52.5, 2 51.2),
    (5 53, 3.5 53, 3.5 52.2, 4 52.2, 4.7 52, 5 51.6, 4.7 51.2, 4 51, 3.5 51.1),
    (6.5 51, 6.5 53, 7.5 53, 8 52.8, 8.2 52.5, 8 52.2, 7.5 52, 6.5 52, 8.2 51)
  )"
)

cells <- a5_linestring_to_cells(a5R_strokes, resolution = 9)
length(cells)

plot(a5_cell_to_boundary(cells),
     col = "#206ead80", border = "#37af6d", asp = 1)
```

Each stroke is one continuous linestring; the function walks the
great-circle path, expanding cell neighbours along the way, and
returns the union across all strokes.

