---
title: "Getting started with a5R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with a5R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(a5R)
```

## Index a point

Map a longitude/latitude coordinate to a cell at a given resolution (0--30).
Higher resolutions produce smaller cells.

```{r}
cell <- a5_lonlat_to_cell(-3.19, 55.95, resolution = 5)
cell
```

Convert back to the cell centre point:

```{r}
a5_cell_to_lonlat(cell)
```

## Cell boundaries

Get the boundary polygon for one or more cells:

```{r boundary-plot, fig.width = 5, fig.height = 4}
boundary <- a5_cell_to_boundary(cell)
boundary

plot(boundary, col = "#206ead20", border = "#206ead", asp = 1)
```

Boundaries are returned as `wk_wkb` vectors by default (set `format = "wkt"` for
WKT). Both integrate directly with sf, terra, and other spatial tooling via the
wk package.

## Hierarchy

Every cell has a parent at a coarser resolution and children at finer
resolutions. A5 cells have 4 children per level.

```{r}
a5_cell_to_parent(cell)
children <- a5_cell_to_children(cell)
children
```

Cell area decreases geometrically with resolution:

```{r}
a5_cell_area(0:5)
```

## Grid generation

`a5_grid()` returns all cells at a target resolution that cover a given area.
Pass a bounding box as a numeric vector:
```{r grid-plot, fig.width = 5, fig.height = 4}
cells <- a5_grid(c(-3.3, 55.9, -3.1, 56.0), resolution = 8)
length(cells)

plot(a5_cell_to_boundary(cells), col = "#206ead20", border = "#206ead", asp = 1)
```

Any geometry that wk can handle works too --- polygons, sf objects, or even
`a5_cell` vectors:

```{r}
poly <- wk::wkt("POLYGON ((-3.3 55.9, -3.1 55.9, -3.1 56, -3.3 56, -3.3 55.9))")
a5_grid(poly, resolution = 5)
```

## Compact and uncompact

When a complete set of siblings is present, `a5_compact()` merges them into
their shared parent, reducing the number of cells without losing coverage:

```{r}
children <- a5_cell_to_children(cell)
length(children)

compacted <- a5_compact(children)
compacted

# Round-trip back to the original set
a5_uncompact(compacted, resolution = 6)
```
