---
title: "Working with Arrow and Parquet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Arrow and Parquet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The uint64 problem

A5 cell IDs are 64-bit unsigned integers. R has no native `uint64` type,
and its `double` can only represent integers exactly up to 2^53. Nearly
half of all A5 cell IDs exceed this threshold, so converting them to
`double` silently corrupts the data.

This is a problem when reading Parquet files that store A5 cell IDs as
`uint64` columns — the standard format used by DuckDB, Python, and
[geoparquet.io](https://geoparquet.io/). By default, `arrow::read_parquet()`
converts `uint64` to R's `double`, losing precision:

```{r naive, eval = requireNamespace("arrow", quietly = TRUE)}
library(arrow)

# Simulate: two uint64 values, one below and one above 2^53
below <- 576460752303423494   # fits exactly in double
above <- 576460752303423494 + 576460752303423494  # exceeds 2^53

cat("2^53:    ", format(2^53, scientific = FALSE), "\n")
cat("'above': ", format(above, scientific = FALSE), "\n")
cat("Same as 'above + 1'?", above == above + 1, "\n")
# TRUE — precision has already been lost
```

## The solution: `a5_cell_from_arrow()` and `a5_cell_to_arrow()`

a5R provides two functions that bypass the lossy `double` conversion
entirely, using Arrow's zero-copy `View()` to reinterpret the raw bytes:

```{r bridge, eval = requireNamespace("arrow", quietly = TRUE)}
library(a5R)
library(tibble)

# Six cities across the globe — some will have bit 63 set (origin >= 6)
cities <- tibble(
  name = c("Edinburgh", "Tokyo", "São Paulo", "Nairobi", "Anchorage", "Sydney"),
  lon  = c(   -3.19,     139.69,     -46.63,     36.82,    -149.90,    151.21),
  lat  = c(   55.95,      35.69,     -23.55,     -1.29,      61.22,    -33.87)
)

cities$cell <- a5_lonlat_to_cell(cities$lon, cities$lat, resolution = 10)
cities
```

These cells work seamlessly in tibbles. Now let's enrich the data with
some A5 operations — cell area, resolution, and distance from Edinburgh:

```{r enrich, eval = requireNamespace("arrow", quietly = TRUE)}
edinburgh <- cities$cell[1]

cities$area_km2 <- as.numeric(a5_cell_area(10)) / 1e6
cities$resolution <- a5_get_resolution(cities$cell)
cities$dist_from_edinburgh_km <- as.numeric(
  a5_cell_distance(cities$cell, rep(edinburgh, nrow(cities)))
) / 1000

cities
```

## Writing and reading Parquet

Convert to an Arrow table and write to Parquet. The cell column is stored
as native `uint64` — the same binary format used by DuckDB, Python, and
geoparquet.io:

```{r parquet_write, eval = requireNamespace("arrow", quietly = TRUE)}
tf <- tempfile(fileext = ".parquet")

arrow_tbl <- arrow::arrow_table(
  name = cities$name,
  cell_id = a5_cell_to_arrow(cities$cell),
  area_km2 = cities$area_km2,
  dist_from_edinburgh_km = cities$dist_from_edinburgh_km
)
arrow_tbl$schema
arrow::write_parquet(arrow_tbl, tf)
```

Read it back — `a5_cell_from_arrow()` recovers the exact cell IDs
without any precision loss:

```{r parquet_read, eval = requireNamespace("arrow", quietly = TRUE)}
pq <- arrow::read_parquet(tf, as_data_frame = FALSE)

# Recover cells from the uint64 column, bind with the rest of the data
recovered_cells <- a5_cell_from_arrow(pq$column(1))
result <- as.data.frame(pq)
result$cell <- recovered_cells
result <- tibble::as_tibble(result[c("name", "cell", "area_km2", "dist_from_edinburgh_km")])
result
```

Verify the round-trip is lossless:

```{r verify, eval = requireNamespace("arrow", quietly = TRUE)}
identical(format(cities$cell), format(result$cell))
```

## How it works under the hood

1. **`a5_cell_to_arrow()`**: packs the eight raw-byte fields into 8-byte
   little-endian blobs (one per cell), creates an Arrow
   `fixed_size_binary(8)` array, then uses `View(uint64)` to
   reinterpret the bytes as unsigned 64-bit integers — zero-copy.

2. **`a5_cell_from_arrow()`**: does the reverse — `View(fixed_size_binary(8))`
   on the `uint64` array to get the raw bytes, then unpacks each 8-byte
   blob into the eight raw-byte fields used by `a5_cell`.

The raw bytes never pass through `double`, so there is no precision loss
at any step. See `vignette("internal-cell-representation")` for details
on the raw-byte representation.
