---
title: "How a5R stores cell IDs without strings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How a5R stores cell IDs without strings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The problem

An A5 cell ID is a 64-bit unsigned integer (`u64`).
R has no native `u64` type — its integers are 32-bit signed (`-2^31` to
`2^31 - 1`), and its doubles are 64-bit IEEE 754 floating point. A `double`
can only represent integers exactly up to 2^53, while a `u64` can go up to
2^64 - 1.

The obvious workaround is to store cell IDs as hex strings
(`"0800000000000006"`). This works, but every trip across the R ↔ Rust
boundary requires hex parsing and formatting — O(n) string allocation that
dominates the cost of lightweight operations like `a5_get_resolution()` or
`a5_cell_to_parent()`.

## The solution: split into two 32-bit halves

A `u64` has exactly 64 bits. We split it into two 32-bit unsigned integers
and store each as an R `double`:

```
cell_id (u64):  0x0800000000000006

hi (upper 32 bits):  0x08000000  →  134217728.0  (stored as R double)
lo (lower 32 bits):  0x00000006  →          6.0  (stored as R double)
```

This is lossless because:

- A `u32` ranges from 0 to 4,294,967,295 (< 2^32)
- An R `double` represents all integers up to 2^53 exactly
- So every `u32` value survives the `u32 → f64 → u32` round-trip perfectly

On the Rust side, the split and recombination are single bitwise operations:

```rust
// u64 → two f64 values
fn u64_to_hilo(id: u64) -> (f64, f64) {
    ((id >> 32) as f64, (id & 0xFFFFFFFF) as f64)
}

// two f64 values → u64
fn hilo_to_u64(hi: f64, lo: f64) -> Option<u64> {
    if hi.is_nan() || lo.is_nan() { return None; }
    Some(((hi as u64) << 32) | (lo as u64))
}
```

No string allocation, no parsing — just bit shifts and casts.

## R-side: a vctrs record type

On the R side, `a5_cell` is a **vctrs record** (`vctrs::new_rcrd()`) with two
fields:

```{r}
library(a5R)
cell <- a5_lonlat_to_cell(-3.19, 55.95, resolution = 10)
vctrs::field(cell, "hi")
vctrs::field(cell, "lo")
```

The two fields are plain numeric vectors — contiguous blocks of memory with
no per-element overhead. Subsetting, combining, and NA propagation are all
handled automatically by vctrs.

Hex strings are only produced on demand:

```{r}
# Display calls format(), which converts to hex for readability
cell

# Explicit conversion
a5_u64_to_hex(cell)

# Round-trip from hex
a5_cell("0800000000000006")
```

## Why this matters

Compare memory for one million cells:

```{r}
set.seed(42)
cells <- a5_lonlat_to_cell(
  runif(1e6, -180, 180),
  runif(1e6, -80, 80),
  resolution = 10
)

# rcrd: two contiguous double vectors (2 × 8 bytes × 1M = 15.3 MB)
format(object.size(cells), units = "MB")

# equivalent hex strings would be ~81 MB
# (16 chars + 56-byte SEXP header per string)
hex <- a5_u64_to_hex(cells)
format(object.size(hex), units = "MB")
```

Operations that were previously bottlenecked by hex parsing — like
`a5_cell_to_parent()` or `a5_compact()` — now run at near-native speed
because the `u64` cell ID is reconstructed from two doubles with a bit shift,
not parsed from a 16-character string.

## NA handling

R represents missing doubles as `NA_real_`, a special "not a number" value.
On the Rust side, `hilo_to_u64` detects this and returns `None`, which maps
back to `NA` in both fields on the R side. This means standard R idioms work
as expected:

```{r}
cells_with_na <- a5_cell(c("0800000000000006", NA))
is.na(cells_with_na)
```

## Summary

| Aspect | Hex strings | Hi/lo doubles |
|--------|------------|---------------|
| R type | `character` vector | `vctrs_rcrd` (two `double` fields) |
| Memory (1M cells) | ~81 MB | ~15 MB |
| R ↔ Rust crossing | O(n) hex parse/format | O(1) bit shift per element |
| Human-readable | Always | On `format()` / `print()` |
| Lossless | Yes | Yes (u32 ⊂ exact doubles) |
