---
title: "How a5R stores cell IDs without strings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How a5R stores cell IDs without strings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The problem

An A5 cell ID is a 64-bit unsigned integer (`u64`).
R has no native `u64` type: its integers are 32-bit signed (`-2^31` to
`2^31 - 1`), and its doubles are 64-bit floating point. A `double`
can only represent integers exactly up to 2^53, while a `u64` can go up to
2^64 - 1.

The obvious workaround is to store cell IDs as hex strings
(`"0800000000000006"`). This works, but every trip across the R / Rust
boundary requires hex parsing and formatting: O(n) string allocation that
dominates the cost of lightweight operations like `a5_get_resolution()` or
`a5_cell_to_parent()`.

## The solution: eight raw-byte fields

A `u64` is exactly 8 bytes. We store each byte of the little-endian
representation as a separate `raw` vector field in a vctrs record type, with
one twist: the most significant byte, `b8`, is stored XOR `0xFC` (see
[NA handling](#na-handling) for why).

```
cell_id (u64):  0x0800000000000006

little-endian bytes:
  0x06, 0x00, 0x00, 0x00, 0x00, 0x00, 0x00, 0x08

stored fields:
  b1 = 0x06, b2 = 0x00, b3 = 0x00, b4 = 0x00,
  b5 = 0x00, b6 = 0x00, b7 = 0x00, b8 = 0x08 XOR 0xFC = 0xF4
```

This is lossless: XOR with a constant is its own inverse, so the eight
stored bytes map one-to-one onto the original `u64`. On the Rust side,
reconstructing the id from the eight byte slices is one XOR and a single
`u64::from_le_bytes()` call. There are no pointers, so an `a5_cell` can be
saved with `saveRDS()` like any other R object. Objects saved by a5R 0.6.0
or earlier used the plain byte layout and do not read correctly in later
versions; re-create them from hex strings or Arrow, which store the true id.

## R-side: a vctrs record type

On the R side, `a5_cell` is a **vctrs record** (`vctrs::new_rcrd()`) with
eight fields (`b1` through `b8`):

```{r}
library(a5R)
cell <- a5_lonlat_to_cell(-3.19, 55.95, resolution = 10)
vctrs::field(cell, "b1")
vctrs::field(cell, "b8") # stored XOR 0xFC
```

Each field is a plain `raw` vector: a contiguous block of memory with
no per-element overhead. Subsetting, combining, and NA propagation are all
handled automatically by vctrs.

Hex strings are only produced on demand:

```{r}
# Display calls format(), which converts to hex for readability
cell

# Explicit conversion
a5_u64_to_hex(cell)

# Round-trip from hex
a5_cell("0800000000000006")
```

## Why this matters

Compare memory for one million cells:

```{r}
set.seed(42)
cells <- a5_lonlat_to_cell(
  runif(1e6, -180, 180),
  runif(1e6, -80, 80),
  resolution = 10
)

# rcrd: eight contiguous raw vectors (8 × 1 byte × 1M ≈ 7.6 MB)
format(object.size(cells), units = "MB")

# equivalent hex strings would be ~81 MB
# (16 chars + 56-byte SEXP header per string)
hex <- a5_u64_to_hex(cells)
format(object.size(hex), units = "MB")
```

## NA handling

`NA` is the id `0xFC00000000000000`. It is never a valid cell: at
resolutions 0 to 29 the top 6 bits hold the quintant (0 to 59), and 63
(binary `111111`) is unused; the id does not decode as a resolution-30 cell
either. Only the full id is the sentinel. Resolution 30 keeps just 5, 3 or 1
quintant bits at the top to make room for position bits, so valid
resolution-30 cells can have a top byte of `0xFC`.

Storing `b8` XOR `0xFC` makes the sentinel's stored form eight `00` bytes.
That matters because vctrs fills gaps field by field, and a missing `raw`
value is `00`. Combining with `NA`, subsetting with an `NA` index, assigning
`NA`, and unmatched rows in joins all produce eight `00` bytes, which decode
to `NA`. With plain bytes they would decode to id 0, the world cell, a valid
id.

On the Rust side, the sentinel is detected and mapped to `None`. Standard
R idioms work as expected:

```{r}
cells_with_na <- a5_cell(c("0800000000000006", NA))
is.na(cells_with_na)
is.na(c(cell, NA))
```

## Working with base R

Because an `a5_cell` is a list of eight raw fields underneath, base
functions that look at that structure directly rather than dispatching on
the class need some care. a5R covers the common cases; the rest have a
vctrs form that is both correct and fast.

**Works as expected.** Subsetting, `c()`, `rev()`, `unique()`,
`duplicated()`, `sort()`, `order()`, `is.na()`, `sample()`, `==`,
comparison operators, `split(cells, f)` with a factor `f`, data frame and
tibble columns, and `rbind()` of data frames with `a5_cell` columns.
Assigning past the end (`x[length(x) + 1] <- cell`) and `length(x) <- n`
grow the vector with `NA`, as they do for base vectors.

**Fast.** `match()` and `%in%` use an exact `mtfrm()` key rather than hex
strings, so they cost about the same as `vctrs::vec_match()` and
`vctrs::vec_in()`.

**Use the vctrs form instead.** Two base functions inspect the record's
fields and cannot be intercepted by a method:

- `split(x, cells)` and `tapply(x, cells, f)` with an `a5_cell` *grouping*
  vector see a list of eight raw vectors and group on their interaction,
  giving one group or an error. Use `vctrs::vec_split(x, cells)` or
  `vctrs::vec_group_id(cells)`, or group with `factor(cells)`.
- `unlist()` on a plain list of `a5_cell` vectors, such as `lapply()`
  output, descends into the fields and returns a meaningless `raw` vector.
  Use `vctrs::list_unchop()`, `do.call(c, x)`, or wrap the list with
  `as_a5_cell_list()`, after which `unlist()` works. Lists returned by
  `a5_cell_to_children()` and friends with `simplify = FALSE` are already
  `a5_cell_list` objects.

Adding rows to a data frame by index (`df[nrow(df) + 1, ] <- ...`) is not
supported for any record-backed column, because base R strips the class
before growing the column. Use `rbind()` or `vctrs::vec_rbind()`.

```{r base-r, eval = FALSE}
cells <- a5_lonlat_to_cell(c(0, 10, 0), c(0, 10, 0), resolution = 5)

# Grouping by cell: use vctrs
vctrs::vec_split(1:3, cells)$val
#> [[1]] 1 3   [[2]] 2

# Flattening lapply() output
parts <- lapply(1:3, function(i) a5_cell_to_children(cells[i]))
unlist(as_a5_cell_list(parts))
```

## Summary

| Aspect | Hex strings | Raw bytes |
|--------|------------|-----------|
| R type | `character` vector | `vctrs_rcrd` (eight `raw` fields) |
| Memory (1M cells) | ~81 MB | ~7.6 MB |
| R-Rust crossing | O(n) hex parse/format | Zero-copy byte access |
| Human-readable | Always | On `format()` / `print()` |
| Lossless | Yes | Yes (exact byte representation) |
