---
title: "Moving between resolutions by location"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Moving between resolutions by location}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(a5R)
```

`a5_cell_to_spatial_parent()` and `a5_cell_to_spatial_children()` are
experimental. Their names and behaviour may change.

## The index hierarchy is not spatially nested

`a5_cell_to_parent()` and `a5_cell_to_children()` follow the A5 index. A
cell's index children do not tile it: some of them lie mostly in its
neighbours, and some cells inside it belong to a neighbour's index.

The spatial functions follow location instead. A fine cell belongs to the
coarse cell containing its centre.

```{r children-plot, fig.width = 7, fig.height = 3.6, echo = -c(1, 11)}
op <- par(mfrow = c(1, 2), mar = c(0.5, 0.5, 2, 0.5))
cell <- a5_lonlat_to_cell(-53.5, -19, resolution = 8)
index_kids <- a5_cell_to_children(cell, resolution = 11)
spatial_kids <- a5_cell_to_spatial_children(cell, resolution = 11)

outline <- a5_cell_to_boundary(cell)
lim <- wk::wk_bbox(a5_cell_to_boundary(c(index_kids, spatial_kids)))
lim <- unclass(lim)

for (panel in list(list(index_kids, "Index children"),
                   list(spatial_kids, "Spatial children"))) {
  plot(NULL, xlim = c(lim$xmin, lim$xmax), ylim = c(lim$ymin, lim$ymax),
       asp = 1, axes = FALSE, xlab = "", ylab = "", main = panel[[2]])
  plot(a5_cell_to_boundary(panel[[1]]), col = "#206ead40",
       border = "#206ead", add = TRUE)
  plot(outline, border = "#333333", lwd = 2, add = TRUE)
}
par(op)
```

Both sets hold 64 cells here, but only the spatial set sits inside the
outline. The mismatch is not rare:

```{r}
set.seed(1)
fine <- a5_lonlat_to_cell(
  stats::runif(1e4, -54, -53), stats::runif(1e4, -19.5, -18.5),
  resolution = 18
)
mean(a5_cell_to_parent(fine, 15) != a5_cell_to_spatial_parent(fine, 15))
```

For about a third of res-18 cells, the index parent at res 15 is a
neighbour of the cell they lie in.

The upward direction shows the same thing from the other side. Here is one
res-10 cell whose index parent at res 8 is not the cell it lies in:

```{r parent-plot, fig.width = 5, fig.height = 4, echo = -c(1, 17)}
op <- par(mar = c(0.5, 0.5, 0.5, 0.5))
kids <- a5_cell_to_children(a5_grid_disk(cell, 1), resolution = 10)
moved <- kids[a5_cell_to_parent(kids, 8) != a5_cell_to_spatial_parent(kids, 8)]
# the clearest case: the fine cell furthest from its index parent's centre
fine_cell <- moved[which.max(a5_cell_distance(moved, a5_cell_to_parent(moved, 8)))]
index_parent <- a5_cell_to_parent(fine_cell, 8)
spatial_parent <- a5_cell_to_spatial_parent(fine_cell, 8)

lim <- unclass(wk::wk_bbox(a5_cell_to_boundary(c(index_parent, spatial_parent))))

plot(NULL, xlim = c(lim$xmin, lim$xmax), ylim = c(lim$ymin, lim$ymax),
     asp = 1, axes = FALSE, xlab = "", ylab = "")
plot(a5_cell_to_boundary(index_parent), col = "#d9534f20", border = "#d9534f",
     lwd = 2, lty = 2, add = TRUE)
plot(a5_cell_to_boundary(spatial_parent), col = "#206ead20", border = "#206ead",
     lwd = 2, add = TRUE)
plot(a5_cell_to_boundary(fine_cell), col = "#333333", border = NA, add = TRUE)
plot(a5_cell_to_lonlat(fine_cell), pch = 19, cex = 0.8, col = "white", add = TRUE)
legend("bottomright", bty = "n", lwd = 2, lty = c(2, 1),
       col = c("#d9534f", "#206ead"), legend = c("Index parent", "Spatial parent"))
par(op)
```

The fine cell (dark, with its centre in white) sits inside the spatial
parent, while its index parent is the neighbouring cell.

## Aggregating to a coarser grid

Take a complete res-18 layer over a small area containing a circular patch,
such as a stand of forest: cells inside the patch have value 1, cells
outside have value 0. Average it to res 15 under each rule.

```{r}
area <- a5_lonlat_to_cell(-53.5, -19, resolution = 12)
layer <- a5_cell_to_children(a5_grid_disk(area, 1), resolution = 18)
length(layer)

centre <- unclass(a5_cell_to_lonlat(area))
in_patch <- function(p) {
  p <- unclass(p)
  as.numeric((p$x - centre$x)^2 + (p$y - centre$y)^2 < 0.004^2)
}
fine <- data.frame(value = in_patch(a5_cell_to_lonlat(layer)))

mean_by <- function(cell, value) {
  stats::aggregate(list(value = value), list(cell = format(cell)), mean)
}
by_index <- mean_by(a5_cell_to_parent(layer, 15), fine$value)
by_spatial <- mean_by(a5_cell_to_spatial_parent(layer, 15), fine$value)
```

For a reference that uses neither hierarchy, drop two million random points
on the area, place each directly in its res-15 cell, and take the share of
points per cell that fall in the patch. This estimates each coarse cell's
true patch fraction.

```{r}
set.seed(1)
bb <- unclass(wk::wk_bbox(a5_cell_to_lonlat(layer)))
pts <- list(x = stats::runif(2e6, bb$xmin, bb$xmax),
            y = stats::runif(2e6, bb$ymin, bb$ymax))
truth <- mean_by(a5_lonlat_to_cell(pts$x, pts$y, resolution = 15), in_patch(pts))
names(truth)[2] <- "truth"

error <- function(agg) {
  m <- merge(agg, truth)
  m <- m[m$value > 0 | m$truth > 0, ]
  round(c(mean = mean(abs(m$value - m$truth)), max = max(abs(m$value - m$truth))), 3)
}
error(by_index)
error(by_spatial)
```

Aggregating by index parent misstates the patch fraction of the cells it
touches by about a tenth on average, and by a third at worst, because each
coarse cell averages fine cells from its neighbours. The spatial rule is off
only by the sampling noise of the reference. `format()` supplies a grouping
key here because base grouping functions cannot group an `a5_cell` vector
directly (see `vignette("internal-cell-representation")`).

## Expanding coarse cells

The downward direction is the exact inverse. Expanding a set of coarse cells
gives each fine cell once, inside the coarse cell whose centre rule claims
it:

```{r}
patch <- a5_uncompact(a5_grid_disk(cell, 1), 8)
kids <- a5_cell_to_spatial_children(patch, resolution = 11, simplify = FALSE)
lengths(kids)

all_kids <- unlist(kids)
vctrs::vec_duplicate_any(all_kids)
all(a5_cell_to_spatial_parent(all_kids, 8) == rep(patch, lengths(kids)))
```

Counts vary slightly from cell to cell but average `4^3 = 64` for these three
resolution steps, as for index children.

## Which to use

Use the index functions for operations on ids: compaction, id ranges and
Hilbert ordering. Use the spatial functions whenever values move between
resolutions and must stay where they are on the ground.

The spatial functions cost more. The index functions are bit operations;
`a5_cell_to_spatial_parent()` locates each centre, taking about as long as
`a5_lonlat_to_cell()`. Both spatial functions run in parallel with
`a5_set_threads()`.
